---
title: "HAT hepatic fat study"
output: github_document
---

```{r setup, include=FALSE}
# Setup -------------------------------------------------------------------

# Required packages
pacs <- c("tidyverse", "haven", "ggExtra", "ggResidpanel")
sapply(pacs, require, character.only = TRUE)

```

## Datasets

* A CSV file `HAT_KeyVariables.csv` includes PIDs, their assigned group and date randomized, n = 1008

* A CSV file `HAT_HFF.csv` includes PID, MRI dates and 3 variables about hepatic fat MRI measurements:
  * `hff`: Hepatic fat fraction
  * `FWHM`: Full width half maximum (the width of the peak measured at 50% of its maximum height), smaller the better
  * `sn`: Signal/noise ratio, higher the better
  * The file is in long format, having up to 2 MRI measurements per subject: n obs = 1932
  
* Demographic data: A SAS data file `final_n961_foodgroups_june2024.sas7bdat`
  * n = 961
  * Includes demographics (age, gender, race, education), BMI, and some dietary data

* Dietary data: A CSV file `HAT_GL_GI_by_FG_updated_080224.csv`
  * n.obs = 2845 from n = 961 distinct PIDs
  * Include glycemic load/index from various food groups by subject and visit
  
## Data check on hepatic fat

```{r read_hff, echo = FALSE, warning = FALSE, message = FALSE}
# Date randomized
# n distinct = 1008
randdate <- read_csv("./data/HAT_KeyVariables.csv") %>% 
  mutate(d_randomized = as.Date(d_randomized, format = "%d%b%Y")) %>% 
  select(pid, d_randomized)

# MRI measurement of hepatic fat
# n obs = 1932
# n distinct = 1008
# hfat <- read_csv("./data/HAT_HFF.csv") %>% 
hfat <- read_csv("./data/HAT_HFF_SN_fixed.csv") %>% 
  mutate(mridate = as.Date(mridate, format = "%d%b%y")) %>% 
  inner_join(randdate, by = "pid") %>% 
  mutate(prepost = ifelse(mridate <= d_randomized, 0, 1))

# Number of MRIs
n_MRI <- hfat %>% 
  group_by(pid) %>% 
  tally()
```

* The hepatic fat data file `HAT_HFF.csv` includes 1932 MRI measurements from 1008 distinct PIDs
  * The file does not have a variable indicating if a measurement is a pre- or post-intervention
  * The date of randomization was added from `HAT_KeyVariables.csv` to determine pre/post measurements

* Most of the subjects (n = 924) have two MRI measurements:
  * There were 3 subjects who have 2 measurements **after** randomization **[Decision needed]**
  
* There are 81 subjects who have only one MRI measurement
  * Among them, 75 subjects have only pre-intervention measurement **[Decision needed]**
  * Six subjects have only post-measurement **[Decision needed]**

* There are 3 subjects who did not have any MRI measurements (i.e., MRI date missing, measurement missing)

## Preliminary analysis on hepatic fat measurements

* Disclaimer: The following analysis was done regardless of pre/post measurements

* Descriptive statistics on `hff`, `FWHM`, `sn`:
  * The HFF value ranges from 0.04% to 54%, with mean of 10% (median 5.7%)
    * Its distribution appears to be very right-skewed
  * ~~Note the negative value of -1 on S/N ratio -- what does this indicate? **[Clarify]**~~
  * ~~There are 7 MRI measurements that have a value `-1` on S/N ratio~~
  * The issue of S/N = -1 has been resolved -- see the email correspondence on 1/27/25

```{r hff_descriptive, echo = FALSE}
hfat %>% 
  select(hff, FWHM, sn) %>% 
  summary()
```

### Distribution of HFF

* A histogram and a cumulative density function of HFF are shown below
  * Vertical dash lines divide HFF values into 5 groups at 5, 10, 20, and 30%
  * The distribution of HFF is highly right-skewed, but very smooth without apparent separating points
  * **[Decision needed]** What should we draw a line for outliers on HFF?

```{r hff_distrib, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 4}
p1 <- hfat %>% 
  ggplot(aes(x = hff)) +
  geom_histogram() +
  geom_vline(xintercept = c(0.05, 0.1, 0.2, 0.3), linetype = 2) +
  labs(x = "Hepatic fat fraction")

p2 <- hfat %>% 
  ggplot(aes(x = hff)) +
  stat_ecdf(geom = "step") +
  geom_vline(xintercept = c(0.05, 0.1, 0.2, 0.3), linetype = 2) +
  scale_y_continuous(n.breaks = 11) +
  labs(x = "Hepatic fat fraction", y = "Empirical CDF")

library(patchwork)
p1 + p2
```

* A frequency table of HFF (5 groups):
  * About 45% of measurements are "normal" having HFF <5%

```{r hff_freqtab, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 4}
hfat %>% 
  filter(!is.na(hff)) %>% 
  mutate(hff_cat = cut(hff, breaks = c(0, 0.05, 0.1, 0.2, 0.3, Inf)),
         hff_cat = factor(hff_cat, labels = c("Normal  : 0-5", "Mild    : 5-10", "Moderate: 10-20", "Severe  : 20-30", "Advanced: >30"))) %>% 
  group_by(hff_cat) %>% 
  tally() %>% 
  mutate(pct = n / sum(n) * 100)
```

### Change in HFF

* For those who have both pre- and post-measurements, changes in HFF value (post - pre) were calculated
* A histogram of change in HFF is shown below
  * Most of subjects have an absolute change <0.2

```{r hff_change, echo = FALSE, warning = FALSE, message = FALSE}
hfat_wide <- hfat %>% 
  filter(!is.na(mridate)) %>% 
  filter(!pid %in% c(15154629, 15217132, 15235033)) %>% 
  mutate(prepost = factor(prepost, labels = c("Pre", "Post"))) %>%
  select(pid, hff, FWHM, prepost) %>% 
  pivot_wider(names_from = prepost, values_from = c(hff, FWHM))

hfat_wide %>% 
  mutate(hff_diff = hff_Post - hff_Pre) %>% 
  ggplot(aes(x = hff_diff)) +
  geom_histogram(aes(y = after_stat(count / sum(count)))) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Change in HFF (post - pre)",
       y = "Percent")
```

* See the list below for those whose HFF value changed >0.2. There are 11 such subjects:
  * **[Clarify]** Is it likely to see such change in HFF during the study period?

```{r hff_change_list, echo = FALSE, warning = FALSE, message = FALSE}
hfat_wide %>% 
  mutate(hff_diff = hff_Post - hff_Pre) %>% 
  filter(abs(hff_diff) > .2) %>% 
  select(pid, hff_Pre, hff_Post, hff_diff)

pid_large_hff_change <- hfat_wide %>% 
  mutate(hff_diff = hff_Post - hff_Pre) %>% 
  filter(abs(hff_diff) > .2) %>% 
  distinct(pid)
```

* There are 23 subjects whose HFF value changed >0.15 (not shown here)


### S/N ratio and FWHM

* A scatterplot of S/N and FWHM is shown below, with histograms on margins.
  * Those measurements of S/N = -1 are shown red in the figure
  * Notice that both distributions are right-skewed
  * **[Decision needed]** Where should we draw lines for potential outliers?
    * For example, there are 18 measurements that have S/N < 100 (more noise) and FWHM > 0.6 (lower resolution)

```{r sn_fwhm_scatter, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 5, fig.height = 5}
p <- hfat %>% 
  ggplot(aes(x = sn, y = FWHM)) +
  geom_point(color = "black") +
  # geom_point(data = hfat[hfat$sn < 0,], aes(x = sn, y = FWHM), color = "red") +
  labs(x = "S/N")

ggMarginal(p, type = "histogram")
```

### Relationship between HFF and BMI

* A scatterplot between BMI and HFF is shown below
  * Both axes are log-scale, due to their skewed distributions
  * A smoothed trend curve is super-imposed (with 95% confidence intervals)

```{r hff_bmi_scatter, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 5, fig.height = 5}
# Subjects: n = 961
demog <- read_sas("./data/final_n961_foodgroups_june2024.sas7bdat")

hfat2 <- hfat %>% 
  left_join(demog %>% select(pid, bmi), by = "pid")

test <- hfat2 %>% 
  # filter(sn >= 100, FWHM <= 0.6) %>% 
  ggplot(aes(x = bmi, y = hff)) +
  geom_point() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  geom_smooth() +
  labs(x = "BMI", y = "HFF")

ggMarginal(test, type = "histogram")
```

* The Pearson correlation coefficient between BMI and HFF (both log-scale) was r = 0.32
  * If MRI measurements of S/N < 100 and FWHM > 0.6 were removed, the correlation slightly reduced to r = 0.31

## Exploratory analysis on HFF vs GL/GI

* Based on dietary data, total GL/GI intake was calculated for each visit and then averaged for each subject
* After merging with demographic data and HFF data, there were n = 909 subjects
  * This excludes those who had 2 MRIs after randomization and those who had only baseline MRIs

* Descriptive statistics on hepatic fat fraction (post-intervention only), GL and GI are shown below:

```{r hff_gl_gi_summary, echo = FALSE, warning = FALSE, message = FALSE}

# Dietary data ------------------------------------------------------------

# Subjects: n = 961
diet <- read_csv("./data/HAT_GL_GI_by_FG_updated_080224.csv")

# GL variables
GL_vars <- names(diet) %>% grep("_GL", ., value = TRUE)

# GI variables
GI_vars <- names(diet) %>% grep("_GI", ., value = TRUE)

# Average GL/GI by PID
glgi <- diet %>% 
  mutate(GL = rowSums(across(all_of(GL_vars)))) %>% 
  mutate(GI = rowSums(across(all_of(GI_vars)))) %>%
  group_by(cpartid) %>% 
  summarize(GL = mean(GL), GI = mean(GI)) %>% 
  rename(pid = cpartid)

# Inner-join yields n = 955 subjects
# There are 46 PIDs whose HFF post is missing
# After removing these, n = 909 subjects
df <- demog %>% 
  inner_join(glgi) %>% 
  inner_join(hfat_wide) %>% 
  filter(!is.na(hff_Post))

df %>% 
  select(hff_Post, GL, GI) %>% 
  summary()
```

* Histograms of HFF, GL and GI are shown below
  * Note that the distribution of HFF is highly right-skewed
  * When HFF is used as the dependent variable, this will be log-transformed
  * **[Clarify]** There are some GL values > 300 and GI values > 3000. Are these values plausible? 
  
```{r hff_gl_gi_histograms, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 15, fig.height = 5}
hist_hff <- df %>% 
  ggplot(aes(x = hff_Post)) + 
  geom_histogram() +
  labs(x = "Hepatic fat fraction (post-intervention)")

hist_gl <- df %>% 
  ggplot(aes(x = GL)) + 
  geom_histogram() +
  labs(x = "Glycemic load")

hist_gi <- df %>% 
  ggplot(aes(x = GI)) + 
  geom_histogram() +
  labs(x = "Glycemic index")

hist_hff + hist_gl + hist_gi
```

* Scatterplots between HFF (post-intervention) and GL/GI are shown below
  * These plots are exploratory and not adjusted for any covariates
  * The y-axis (HFF) is on the log scale
  * A smoothed trend is overlaid for each plot
    * Please ignore the tail region of GL/GI where data are sparse and the confidence interval is wide
    * No apparent relationship with HFF (again, unadjusted)
  
```{r hff_gl_gi_scatter, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 5}
scatter_HFF_GL <- df %>% 
  ggplot(aes(x = GL, y = hff_Post)) +
  geom_point() +
  geom_smooth(span = 0.9) +
  scale_y_continuous(trans = "log10") +
  labs(x = "Glycemic load", y = "Hepatic fat fraction (log-scale)") +
  theme(legend.position = "bottom")

scatter_HFF_GI <- df %>% 
  ggplot(aes(x = GI, y = hff_Post)) +
  geom_point() +
  geom_smooth(span = 0.95) +
  scale_y_continuous(trans = "log10") +
  labs(x = "Glycemic index", y = "Hepatic fat fraction (log-scale)") +
  theme(legend.position = "bottom")

scatter_HFF_GL + scatter_HFF_GI
```

## Notes

* Zoom meeting on 1/23/2025 (KL/GS/CH/KO)
  * ~~S/N = -1 indicates invalid values?~~ 
    * ~~Kristie can look up PIDs of those subjects~~
    * ~~Dr. Barnes can look into those images to check if HFF values make sense~~
    * **[Updated]** The issue of S/N = -1 has been resolved -- see the email correspondence on 1/27/25
  
  * It is plausible to see changes of HFF > 0.2 according to Dr. Barnes
    * Dig into past literature with similar intervention as well?
  
  * Any data on compliance on those with a large HFF change during study?
  
  * Need to determine cut-off values for HFF/FWHM/SN outliers

  * Possible covariates to be included in the model
    * Demographics (age/gender/race/education)
    * BMI (pre/post)
    * Any lifestyle variables? -- self-reported
    * Nutrient intake? -- total kcal, SFA (kcal base), sugar?, CHO?
      * Want to identify those overeating

  * KO to proceed analysis without excluding potential outliers for now -- Can exclude any outliers later
  * KL to communicate with Dr. Barnes
  * GS to look into nutrient data
  

